<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{clonotypeR User's Guide}
\usepackage[utf8]{inputenc}
-->

clonotypeR: Identify and analyse B and T cell receptors at a high throughput
============================================================================

    Author: Charles Plessy <plessy@riken.jp>
    Date:   1 May 2013

*clonotypeR* is a R package and accompanying scripts to identify and
analyse clonotypes from high-throughput T cell receptors sequence
libraries. *clonotypeR* is suited to process and organise very large
number of clonotypes, in the order of millions, typically produced by
Roche 454 instruments, and to prepare these sequences for differential
expression analysis with the typical transcriptomics tools as well as
for statistical analysis using existing R packages.

The home page of *clonotypeR* is <http://clonotyper.branchable.com/>.


clonotypeR's workflow
---------------------

Typically, the user receives the output of a next-generation sequencer and runs 
some shell commands that are not part of the *clonotypeR* R package, but that are 
distributed with it on <http://clonotyper.branchable.com/>.

Currently, *clonotypeR* provides only support for mouse sequences, by providing
pre-formatted sequences from `V` and `J` segments.  Support of human sequences
is in preparation.  See <http://clonotyper.branchable.com/references/README>
for more information.

The workflow presented here summarises the different commands to run.  Other
examples are available on line at
<http://clonotyper.branchable.com/doc/workflow/>.

*This example analysis assumes a
[unix](https://en.wikipedia.org/wiki/Operating%5Fsystem%23UNIX%5Fand%5FUNIX%2Dlike%5Foperating%5Fsystems)
system (Linux, Mac OS, …)*


Test data
---------

ClonotypeR come with test data in the `R` package.  While it is not deep enough
to discuss about biology, you can use it to familiarise yourself with the
commands or test them.

The R package is loaded as usual.

```{r load_package}
library(clonotypeR)
````

The data is a table of 120 clonotypes in the `extdata` folder of the package.
The command `read_clonotypes` will parse it in a data frame.  The clonotypes
are arbitrarly assigned to three libraries called `A`, `B`, and `C`.  The
`read_clonotypes` comments determines at load time if the peptidic sequence has
a stop codon or is frame-shifted, and records the information in the
`unproductive` column.

```{r load_data}
clonotypes <- read_clonotypes(system.file('extdata', 'clonotypes.txt.gz', package = "clonotypeR"))
summary(clonotypes)
```

The `clonotype_table` command counts how many times a given clonotype is found
in each library.  It can also count simpler features, in particular `V` and `J`
segments, or any combination of them.

```{r clonotype_table}
head(clonotype_table(levels(clonotypes$lib), data=clonotypes))
head(clonotype_table(levels(clonotypes$lib), "V", data=clonotypes))
head(clonotype_table(levels(clonotypes$lib), "J", data=clonotypes))
```

ClonotypeR provides other functions for further analysis.  `yassai_identifier`
calculates a unique indentifier using the `V`, `J`, peptidic and nucleotidic
information, following the work of [Yassai et al](http://dx.doi.org/10.1007/s00251-009-0383-x).

```{r yassai_identifier}
# Unique identifier
head(yassai_identifier(clonotypes))
```

`unique_clonotypes` and `common_clonotypes` are typically used when comparing libraries.

```{r other_functions}
clonotypes <- clonotype_table(levels(clonotypes$lib), data=clonotypes)

# First six clonotypes of library C
head(unique_clonotypes("C", data=clonotypes))
# Count clonotypes found in library A, and B or C.
length(common_clonotypes(group1="A", group2=c("B","C"), data=clonotypes))
# Matrix of numbers of common clonotypes
common_clonotypes(data=clonotypes)
```

With deeper data, a typical follow-up would be to identify differentially
represnted clonotypes between libraries, for instance with the `edgeR` package,
or to calculate distance between libraries, for instance with the
`vegan` package.


Example data (3 × 2,000 reads)
------------------------------

The data provided on-line at <http://clonotyper.branchable.com/example_data/>
is a sub-sample of three sequence librairies of mouse T cell receptors α (2,000
reads each) made on the 454 Titanium or the 454 junior platforms.  The original
libraries will be deposited in public databanks after publication in a
peer-reviewed journal.

These example libraries are called `A`, `B` and `C`, and are in FASTQ format,
with entries like the following (the sequence was truncated for the convenience
of the display).

    @HKTLYLP01B0MTM
    gactGTCCATCTTCCTTTTATCGGACACTGAAGTATGGATATCAGAAGTGCAgggccttcccacgggaacg
    +
    IIIIIIIIIIIHHFF::::G&gt;IIIGGGIIIIIIIIIGGIIIIIIFEBDCDC&lt;//-5522------


### Detection of V segments

Run the command `clonotyper detect A.fastq` in the same directory as a copy of
the file `A.fastq`.

The result is stored in a temporary directory called `extraction_files`, that
will be created if it does not already exist.

`clonotyper detect` compares the sequences to the reference V segments using
BWA, and produces output like the following.

    [bsw2_aln] read 2000 sequences/pairs (843395 bp)...
    [samopen] SAM header is present: 167 sequences.
    [main] Version: 0.6.2-r126
    [main] CMD: bwa bwasw -t8 /usr/share/clonotypeR/references/V/index A.fastq
    [main] Real time: 1.099 sec; CPU: 8.225 sec

This indicates that 2,000 reads have been processed, representing 843,395 base
pairs in total.  There were 167 reference V segments, and the version number of
BWA was `0.6.2-r126`.  The whole process took less than 10 seconds.

Process the example libraries `B` and `C` similarly with the commands
`clonotyper detect C.fastq` and `clonotyper detect C.fastq`.


### Extraction of CDR3 regions

Run the command `clonotyper extract A` in the same directory as where you ran
`clonotyper detect A.fastq`.  The result is a table stored in a directory
called `clonotypes`, that will be created if it does not already exist.

The output is quite voluminous, and indicates which *V* / *J* combinations are
being found, like on the following.

    TRAV14-3	233
    	TRAJ61	0
    	TRAJ60	0
    	TRAJ59	0
    	TRAJ58	1
    	TRAJ57	39
    	TRAJ56	2
    	TRAJ55	0

The format of the table is explained in the manual page of the function
`read_clonotypes()` of the R package.

For each library (`A`, `B` and `C`), one file is available in the `clonotypes`
directory.  With BWA `0.6.2-r126`, the following numbers of clonotypes are found.

      1072 clonotypes/A.tsv
       924 clonotypes/B.tsv
       689 clonotypes/C.tsv

The files need to be concatenated before analysis in `R`, with the following command.

    find clonotypes/ -name '*tsv' | xargs cat > clonotypes.tsv

A copy of the result file is provided in `inst/extdata/clonotypes2.tsv.xz` for convenience.

### Data analysis in R

Load the clonotypeR library:

```
library(clonotypeR)
```

Load the data in a R object called *clonotypes*:

```
clonotypes <- read_clonotypes('clonotypes.tsv')
```

Alternatively, you can load the convenience copy from
`inst/extdata/clonotypes2.tsv.xz` (see above).

```{r load_data2}
clonotypes <- read_clonotypes(system.file("extdata", "clonotypes2.tsv.xz", package = "clonotypeR"))
```

The command `summary(clonotypes)` already provides useful information.

```{r summary_2}
summary(clonotypes)
```

Identify unique clonotypes, count their sequences in the libraries `A`, `B` and `C`, and store the result as a table arbitrarly named `abc`. 

```{r explore_data2}
abc <- clonotype_table(c('A','B','C'), data=clonotypes)
head(abc)
summary(abc)
```

The summary shows that the most frequent clonotype is in `C`. Using `R` [index vectors](http://cran.r-project.org/doc/manuals/R-intro.html#Index-vectors), we can see that its CDR3 sequence is AASDSNNRIF and that it was not found in the other libraries.

```{r explore_data2b}
abc[abc$C == 124, ]
```

The `clonotype_table` function can also produce a count table for and combination of *V*, *CDR3* or *J* segments.

```{r explore_data2v}
clonotype_table(c('A','B','C'), "V", data=clonotypes)
head(clonotype_table(c('A','B','C'), c("V","J"), data=clonotypes))
```
